### 2024-08-23 machine-learning.qmd
20240823 - change imputation procedure from missForest to missRanger for speed and out-of-sample prediction

```{r}
library("petersenlab")
library("parallel")
library("doParallel")
library("missForest")
library("powerjoin")
library("caret")
library("gpboost")
library("tidyverse")
```

@@

Here is a vignette demonstrating how to impute missing data using `missForest()`: <https://rpubs.com/lmorgan95/MissForest> (archived at: <https://perma.cc/6GB4-2E22>).
Below, we impute the training data and new data separately by position.

@@

```{r}
num_cores

cl <- parallel::makeCluster(num_cores)
doParallel::registerDoParallel(cl)

# QBs
seasonalData_lag_qb_all_imp <- missForest::missForest(
  seasonalData_lag_qb_all %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_qb_all_imp$OOBerror
data_all_qb <- cbind(seasonalData_lag_qb_all[,"gsis_id"], seasonalData_lag_qb_all_imp$ximp)
data_all_qb_matrix <- data_all_qb %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()
newData_qb <- data_all_qb %>% 
  filter(season == max(season, na.rm = TRUE)) %>% 
  select(-fantasyPoints_lag)
newData_qb_matrix <- data_all_qb_matrix[
  data_all_qb_matrix[, "season"] == max(data_all_qb_matrix[, "season"], na.rm = TRUE), # keep only rows with the most recent season
  , # all columns
  drop = FALSE]

seasonalData_lag_qb_train_imp <- missForest::missForest(
  seasonalData_lag_qb_train %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_qb_train_imp$OOBerror
data_train_qb <- cbind(seasonalData_lag_qb_train[,"gsis_id"], seasonalData_lag_qb_train_imp$ximp)
data_train_qb_matrix <- data_train_qb %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()

seasonalData_lag_qb_test_imp <- missForest::missForest(
  seasonalData_lag_qb_test %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_qb_test_imp$OOBerror
data_test_qb <- cbind(seasonalData_lag_qb_test[,"gsis_id"], seasonalData_lag_qb_test_imp$ximp)
data_test_qb_matrix <- data_test_qb %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()

# RBs
seasonalData_lag_rb_all_imp <- missForest::missForest(
  seasonalData_lag_rb_all %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_rb_all_imp$OOBerror
data_all_rb <- cbind(seasonalData_lag_rb_all[,"gsis_id"], seasonalData_lag_rb_all_imp$ximp)
data_all_rb_matrix <- data_all_rb %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()
newData_rb <- data_all_rb %>% 
  filter(season == max(season, na.rm = TRUE)) %>% 
  select(-fantasyPoints_lag)
newData_rb_matrix <- data_all_rb_matrix[
  data_all_rb_matrix[, "season"] == max(data_all_rb_matrix[, "season"], na.rm = TRUE), # keep only rows with the most recent season
  , # all columns
  drop = FALSE]

seasonalData_lag_rb_train_imp <- missForest::missForest(
  seasonalData_lag_rb_train %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_rb_train_imp$OOBerror
data_train_rb <- cbind(seasonalData_lag_rb_train[,"gsis_id"], seasonalData_lag_rb_train_imp$ximp)
data_train_rb_matrix <- data_train_rb %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()

seasonalData_lag_rb_test_imp <- missForest::missForest(
  seasonalData_lag_rb_test %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_rb_test_imp$OOBerror
data_test_rb <- cbind(seasonalData_lag_rb_test[,"gsis_id"], seasonalData_lag_rb_test_imp$ximp)
data_test_rb_matrix <- data_test_rb %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()

# WRs
seasonalData_lag_wr_all_imp <- missForest::missForest(
  seasonalData_lag_wr_all %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_wr_all_imp$OOBerror
data_all_wr <- cbind(seasonalData_lag_wr_all[,"gsis_id"], seasonalData_lag_wr_all_imp$ximp)
data_all_wr_matrix <- data_all_wr %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()
newData_wr <- data_all_wr %>% 
  filter(season == max(season, na.rm = TRUE)) %>% 
  select(-fantasyPoints_lag)
newData_wr_matrix <- data_all_wr_matrix[
  data_all_wr_matrix[, "season"] == max(data_all_wr_matrix[, "season"], na.rm = TRUE), # keep only rows with the most recent season
  , # all columns
  drop = FALSE]

seasonalData_lag_wr_train_imp <- missForest::missForest(
  seasonalData_lag_wr_train %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_wr_train_imp$OOBerror
data_train_wr <- cbind(seasonalData_lag_wr_train[,"gsis_id"], seasonalData_lag_wr_train_imp$ximp)
data_train_wr_matrix <- data_train_wr %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()

seasonalData_lag_wr_test_imp <- missForest::missForest(
  seasonalData_lag_wr_test %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_wr_test_imp$OOBerror
data_test_wr <- cbind(seasonalData_lag_wr_test[,"gsis_id"], seasonalData_lag_wr_test_imp$ximp)
data_test_wr_matrix <- data_test_wr %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()

# TEs
seasonalData_lag_te_all_imp <- missForest::missForest(
  seasonalData_lag_te_all %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_te_all_imp$OOBerror
data_all_te <- cbind(seasonalData_lag_te_all[,"gsis_id"], seasonalData_lag_te_all_imp$ximp)
data_all_te_matrix <- data_all_te %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()
newData_te <- data_all_te %>% 
  filter(season == max(season, na.rm = TRUE)) %>% 
  select(-fantasyPoints_lag)
newData_te_matrix <- data_all_te_matrix[
  data_all_te_matrix[, "season"] == max(data_all_te_matrix[, "season"], na.rm = TRUE), # keep only rows with the most recent season
  , # all columns
  drop = FALSE]

seasonalData_lag_te_train_imp <- missForest::missForest(
  seasonalData_lag_te_train %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_te_train_imp$OOBerror
data_train_te <- cbind(seasonalData_lag_te_train[,"gsis_id"], seasonalData_lag_te_train_imp$ximp)
data_train_te_matrix <- data_train_te %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()

seasonalData_lag_te_test_imp <- missForest::missForest(
  seasonalData_lag_te_test %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_te_test_imp$OOBerror
data_test_te <- cbind(seasonalData_lag_te_test[,"gsis_id"], seasonalData_lag_te_test_imp$ximp)
data_test_te_matrix <- data_test_te %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()

stopCluster(cl)
```

>>>>

```{r}
library("petersenlab")
library("parallel")
library("doParallel")
library("missRanger")
library("powerjoin")
library("caret")
library("gpboost")
library("tidyverse")
```

@@

Here is a vignette demonstrating how to impute missing data using `missForest()`: <https://rpubs.com/lmorgan95/MissForest> (archived at: <https://perma.cc/6GB4-2E22>).
Below, we impute the training data (and all data) separately by position.
We then use the imputed training data to make out-of-sample predictions to fill in the missing data for the testing data.
We do not want to impute the training and testing data together so that we can keep them separate for the purposes of cross-validation.
However, we impute all data (training and test data together) for purposes of making out-of-sample predictions from the machine learning models to predict players' performance next season (when actuals are not yet available for evaluating their accuracy).

```{r}
# QBs
seasonalData_lag_qb_all_imp <- missRanger::missRanger(
  seasonalData_lag_qb_all,
  pmm.k = 5,
  verbose = 2,
  seed = 52242,
  keep_forests = TRUE)

seasonalData_lag_qb_all_imp

data_all_qb <- seasonalData_lag_qb_all_imp$data
data_all_qb_matrix <- data_all_qb %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()
newData_qb <- data_all_qb %>% 
  filter(season == max(season, na.rm = TRUE)) %>% 
  select(-fantasyPoints_lag)
newData_qb_matrix <- data_all_qb_matrix[
  data_all_qb_matrix[, "season"] == max(data_all_qb_matrix[, "season"], na.rm = TRUE), # keep only rows with the most recent season
  , # all columns
  drop = FALSE]

seasonalData_lag_qb_train_imp <- missRanger::missRanger(
  seasonalData_lag_qb_train,
  pmm.k = 5,
  verbose = 2,
  seed = 52242,
  keep_forests = TRUE)

seasonalData_lag_qb_train_imp

data_train_qb <- seasonalData_lag_qb_train_imp$data
data_train_qb_matrix <- data_train_qb %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()

seasonalData_lag_qb_test_imp <- predict(
  object = seasonalData_lag_qb_train_imp,
  newdata = seasonalData_lag_qb_test,
  seed = 52242)

data_test_qb <- seasonalData_lag_qb_test_imp
data_test_qb_matrix <- data_test_qb %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()

# RBs
seasonalData_lag_rb_all_imp <- missRanger::missRanger(
  seasonalData_lag_rb_all,
  pmm.k = 5,
  verbose = 2,
  seed = 52242,
  keep_forests = TRUE)

seasonalData_lag_rb_all_imp

data_all_rb <- seasonalData_lag_rb_all_imp$data
data_all_rb_matrix <- data_all_rb %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()
newData_rb <- data_all_rb %>% 
  filter(season == max(season, na.rm = TRUE)) %>% 
  select(-fantasyPoints_lag)
newData_rb_matrix <- data_all_rb_matrix[
  data_all_rb_matrix[, "season"] == max(data_all_rb_matrix[, "season"], na.rm = TRUE), # keep only rows with the most recent season
  , # all columns
  drop = FALSE]

seasonalData_lag_rb_train_imp <- missRanger::missRanger(
  seasonalData_lag_rb_train,
  pmm.k = 5,
  verbose = 2,
  seed = 52242,
  keep_forests = TRUE)

seasonalData_lag_rb_train_imp

data_train_rb <- seasonalData_lag_rb_train_imp$data
data_train_rb_matrix <- data_train_rb %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()

seasonalData_lag_rb_test_imp <- predict(
  object = seasonalData_lag_rb_train_imp,
  newdata = seasonalData_lag_rb_test,
  seed = 52242)

data_test_rb <- seasonalData_lag_rb_test_imp
data_test_rb_matrix <- data_test_rb %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()

# WRs
seasonalData_lag_wr_all_imp <- missRanger::missRanger(
  seasonalData_lag_wr_all,
  pmm.k = 5,
  verbose = 2,
  seed = 52242,
  keep_forests = TRUE)

seasonalData_lag_wr_all_imp

data_all_wr <- seasonalData_lag_wr_all_imp$data
data_all_wr_matrix <- data_all_wr %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()
newData_wr <- data_all_wr %>% 
  filter(season == max(season, na.rm = TRUE)) %>% 
  select(-fantasyPoints_lag)
newData_wr_matrix <- data_all_wr_matrix[
  data_all_wr_matrix[, "season"] == max(data_all_wr_matrix[, "season"], na.rm = TRUE), # keep only rows with the most recent season
  , # all columns
  drop = FALSE]

seasonalData_lag_wr_train_imp <- missRanger::missRanger(
  seasonalData_lag_wr_train,
  pmm.k = 5,
  verbose = 2,
  seed = 52242,
  keep_forests = TRUE)

seasonalData_lag_wr_train_imp

data_train_wr <- seasonalData_lag_wr_train_imp$data
data_train_wr_matrix <- data_train_wr %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()

seasonalData_lag_wr_test_imp <- predict(
  object = seasonalData_lag_wr_train_imp,
  newdata = seasonalData_lag_wr_test,
  seed = 52242)

data_test_wr <- seasonalData_lag_wr_test_imp
data_test_wr_matrix <- data_test_wr %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()

# TEs
seasonalData_lag_te_all_imp <- missRanger::missRanger(
  seasonalData_lag_te_all,
  pmm.k = 5,
  verbose = 2,
  seed = 52242,
  keep_forests = TRUE)

seasonalData_lag_te_all_imp

data_all_te <- seasonalData_lag_te_all_imp$data
data_all_te_matrix <- data_all_te %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()
newData_te <- data_all_te %>% 
  filter(season == max(season, na.rm = TRUE)) %>% 
  select(-fantasyPoints_lag)
newData_te_matrix <- data_all_te_matrix[
  data_all_te_matrix[, "season"] == max(data_all_te_matrix[, "season"], na.rm = TRUE), # keep only rows with the most recent season
  , # all columns
  drop = FALSE]

seasonalData_lag_te_train_imp <- missRanger::missRanger(
  seasonalData_lag_te_train,
  pmm.k = 5,
  verbose = 2,
  seed = 52242,
  keep_forests = TRUE)

seasonalData_lag_te_train_imp

data_train_te <- seasonalData_lag_te_train_imp$data
data_train_te_matrix <- data_train_te %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()

seasonalData_lag_te_test_imp <- predict(
  object = seasonalData_lag_te_train_imp,
  newdata = seasonalData_lag_te_test,
  seed = 52242)

data_test_te <- seasonalData_lag_te_test_imp
data_test_te_matrix <- data_test_te %>%
  mutate(across(where(is.factor), ~ as.numeric(as.integer(.)))) %>% 
  as.matrix()
```

## Identify Cores for Parallel Processing {#sec-coresParallel}

@@

```{r}
num_cores
```
